Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2016-11-08 manuscript.tex
latest

Evolutionary models differ greatly in their treatment of fitness. In models of genetic evolution, genotypes are typically assigned constant (or frequency-dependent) selection coefficients describing the change in their relative frequencies over time due to differences in viability. This considerably simplifies the mathematics of selection, facilitating greater genetic realism, and can be justified over sufficiently short time intervals \citep[p. 276]{ewens_2012}. However, selection can have very different effects when operating on different types of traits, and evolutionary changes in one population can lead to complicated ecological responses.

@@

Far less work has been done to model fitness in more general terms than particular traits or ecological scenarios, while still capturing key distinctions between different forms of selection. Perhaps this is not surprising given that fitness is such a complex quantity, dependent on all of a phenotype's functional traits \citep{violle_2007} as well as its biotic and abiotic environment. In most cases, a detailed, trait-based, predictive model of fitness would be enormously complicated and have narrow applicability. It is therefore easy to doubt the feasibility of a simplified, general mathematical treatment of fitness \citep[p. 276]{ewens_2012}. Even MacArthur's famous r/K selection scheme is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}. In spite of the r/K scheme's original purpose as an extension of the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}, comparatively few attempts have been made to develop it further as a mathematical analysis of the major different forms of selection.

@@

Trait classification schemes show empirically that, beneath the complicated details of trait variation, even among closely-related species, fitness is predominantly determined by a few key factors such as intrinsic reproducive rate or stress-tolerance. However, while trait classification schemes are firmly grounded in trait data, they are verbal and descriptive rather than mathematical, a recognized hinderance to their broader applicability (e.g. \citep{tilman_2007}).

@@

We broadly follow the spirit of MacArthur's $r$/$K$ selection scheme in that our model is intended to account for fundamentally different forms of selection without getting  entangled in the intricacies of particular ecological scenarios. However, rather than building directly on MacArthur's formalism and its later extensions using Lotka-Volterra equations to incorporate competition (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001}, our model is devised primarily with Grime's C/S/R scheme in mind, and represents a quantitative formalization of how C/S/R manifests at the level of genotype evolution (as opposed to divergence between species). This choice is motivated in part by the substantial empirical support for the C/S/R scheme, and in part by the failings of the $r$/$K$ low/high density dichotomy --- many growth ability traits will confer advantages at both low and high densities (more details in the Discussion).

In section ``Model'' we introduce the basic assumptions of our generalized lottery model. Analytical expressions for the change in genotype abundances over time are introduced in section ``Mean field approximation'', with mathematical details relegated to the Appendices. The following two sections discuss the behaviour of rare mutants and our treatment of Grime's triangle.

@@

Each iteration, adults produce $m_i$ new offspring (henceforth called ``propagules'') which disperse at random over the $U$ unnocupied territories (no dispersal limitation). We assume adults cannot be ousted from occupied territories, so only propagules landing on occupied territories are included in $m_i$. More generally, $m_i$ does not include propagules which never even begin the development cycle. For simplicity, we assume $m_i=b_i n_i$, where $b_i$ is a constant, genotype-specific birth rate.

The number of individuals of the $i$'th genotype landing in any particular territory is denoted $x_i$. Random dispersal implies that in the limit $K\rightarrow \infty$, with $n_i/K$ held fixed, $x_i$ is Poisson distributed with mean territorial propagule density $l_i=m_i/U$. Although $T$ is finite in our model, we assume that $K$ and the $n_i$ are large enough that $x_i$ is Poisson-distributed to a good approximation (details in Appendix A). This dispersal Poisson distribution is denoted $p_i(x_i)$. Note that the large $n_i$, large $T$ approximation places no restrictions on our densities $n_i/T$, but it does preclude consideration of demographic stochasticity when $n_i$ itself is very small (this will be discussed further in Section \ref{sec:invas}).

@@

The increase in $n_i$ over one iteration due to territorial acquisition, $\Delta_+ n_i$, is the sum of genotype $i$'s victories over all $U$ unoccupied territories. Since $p_1(x_1)\ldots p_G(x_G)$ is equal to the proportion of unoccupied territories with $x_1,\ldots,x_G$ of the respective propagules (again, we assume that $T$ is large enough that fluctuations in this proportion are negligible), this sum can be replaced by an expectation over the $p_i$. This gives
\begin{equation}
\Delta_+ n_i(t)=U(t)\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G). \label{eq:growthsumuncoupled}
\end{equation}

In addition to propagule birth and competition, occupied territories become unoccupied due to mortality. For the vast majority of this manuscript we assume that mortality only occurs in adults, and at a constant, genotype-specific per-capita rate $d_i$, so that the overall change in genotype abundances is
\begin{equation}
\Delta n_i(t)=\Delta_+ n_i(t)-d_i n_i(t). \label{eq:delttot}
\end{equation}
We will introduce a different mortality model when we consider the effects of disturbances (Section \ref{sec:grime}), which will also affect competing juveniles.

Note that the competitive ability coefficients $c_i$ represent a strictly relative aspect of fitness in the sense that they only influence population size $N$ indirectly by changing genotype frequencies; that may in turn change the population mean birth and death rates. This can be seen by summing Eq. \eqref{eq:delttot} over genotypes to get the  change in population size $N$, 
\begin{equation}
\Delta N=U(1-e^{-L})-\sum_i d_i n_i,\label{eq:deltN}
\end{equation}
which is independent of $c_i$ (here $L=\sum_j l_j$ is the overall propagule density).

@@

Our mean field approximation is similar to the high-$l_i$ approximation leading to Eq. \eqref{eq:lottery} in that we replace the $x_i$ with appropriate mean values. The key distinction is that territories with a single propagule from the focal genotype, which are critical at low densities, are handled separately. In place of the requirement of $l_i\gg 1$ for all $i$, our approximation only requires that there are no large discrepancies in competitive ability (specifically, that we do not have $c_i/c_j\gg 1$ for any two genotypes; further discussion in section ``Discussion''). We obtain (details in Appendix B)
\begin{equation}
\Delta_+ n_i(t)\approx b_i n_i\left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right], \label{eq:master}
\end{equation}
where
\begin{equation}
R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}\frac{\overline{c}L- c_il_i}{L-l_i}},\label{eq:Dr}
\end{equation}
and
\begin{equation}
A_i=\frac{\overline{c}(1-e^{-l_i})}{c_il_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}+\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right)\sum_{j\neq i}\frac{c_jl_j}{L-l_j}}.\label{eq:Da}
\end{equation}

Comparing Eq. \eqref{eq:master} to Eq. \eqref{eq:lottery}, the classic lottery per-propagule success rate $c_i/\overline{c}L$ has been replaced by three separate terms. The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these territories are won without contest. The second term, $R_i c_i/\overline{c}$ represents competitive victories when the $i$ genotype is a rare invader in a high density population: from Eq. \eqref{eq:Dr}, $R_i\rightarrow 0$ when the $i$ genotype is abundant ($l_i\gg 1$), or other genotypes are collectively rare ($L-l_i\ll 1$). The third term, $A_ic_i/\overline{c}$, represents competitive victories when the $i$ genotype is abundant: $A_i\rightarrow 0$ if $l_i\ll 1$. The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $l_i/L$. If $l_i\gg 1$ for all genotypes, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$). Thus, Eq. \eqref{eq:master} generalizes the classic lottery model to account for arbitrary propagule densities for each genotype.

@@

\begin{figure}
\centering
\includegraphics[scale=0.7]{simulationcomparison.pdf}
\caption{\label{fig:simcomp} The change in genotype abundances in a density dependent lottery model is closely approximated by Eq. \eqref{eq:master}. $\Delta_+ n_i/m_i$ from Eq. \eqref{eq:master} (and its separate components) are shown, along with direct simulations of random dispersal and lottery competition over one iteration over a range of propagule densities (varied by changing $U$ with the $m_i$ fixed). Two genotypes are present. (a) and (b) show low-frequency genotype with $c$-advantage ($m_1/M=0.1$, $c_1=1.5$), (c) and (d) show the high-frequency predominant genotype ($m_2/M=0.9$, $c_2=1$).} 
\end{figure}

@@

To determine how $b$, $c$ and $d$ will evolve in a population where those traits are being modified by mutations, we need to know whether mutant lineages will grow (or decline) starting from low densities. In this section we discuss the behavior of rare genotypes predicted by Eq. \eqref{eq:master}.

Suppose that a population with a single genotype $i$ is in equilibrium. Then $R_i=0$, $\overline{c}=c_i$ and $\Delta n_i = 0$, and so Eq. \eqref{eq:master} gives
\begin{equation}
b_i\left(e^{-L}+A_i\right)-d_i=0.\label{eq:equil}
\end{equation}
Now suppose that a new genotype $j$, which is initially rare, appears in the population. Then $A_j\ll 1$, $l_j\ll L$ and $\overline{c}\approx c_i$, and so, from Eq. \eqref{eq:master}, $n_j$ will increase if 
\begin{equation}
b_j \left(e^{-L}+R_j\frac{c_j}{c_i}\right)-d_j>0.\label{eq:invad}
\end{equation}

Combining Eqs. \eqref{eq:equil} and \eqref{eq:invad}, it is easily verified that if $j$ is superior in one trait, but otherwise identical to $i$, it will invade. Moreover, $j$ will eventually exclude $i$, since it is strictly superior. However, stable coexistence is possible between genotypes that are superior in different traits. To illustrate, suppose that $j$ is better at securing territories ($c_j>c_i$), that $i$ is better at producing propagules ($b_i>b_j$), and that $d_i=d_j$. Coexistence occurs if $j$ will invade an $i$-dominated population, but $i$ will also invade a $j$-dominated population (``mutual invasion''). It is not hard to show that this is possible, since if $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant, then, combining Eqs. \eqref{eq:equil} and \eqref{eq:invad}, we find that $i$ invades $j$ because $b_i>b_j$, while $j$ invades $i$ provided that
\begin{equation}
b_jc_jR_j-b_i c_i A_i>0. \label{eq:jinvadcoex}
\end{equation}
Thus, coexistence occurs if $c_j$ is large enough. Intuitively, the mechanism for coexistence is that territorial contests are important in an $i$-dominated population (high $L$), ensuring that the $c$-specialist $j$ is not excluded, yet territorial contests are irrelevant in a $j$-dominated population (low $L$), ensuring that the $b$-specialist $i$ is not excluded. Fig.~\ref{fig:coex} shows an example of this coexistence between $b$ and $c$ specialists.

@@

If the rare genotype $j$ arises due to mutation, then it's initial low-density behavior is more complicated than the above invasion analysis suggests. The mutant lineage starts with one individual $n_j=1$, and remains at low abundance for many generations after its initial appearance. During this period, the mutant abundance $n_j$ will behave stochastically, and the deterministic equations \eqref{eq:growthsumuncoupled} and \eqref{eq:master} do not apply (Section \ref{sec:model}). However, if $n_j$ becomes large enough, its behavior will become effectively deterministic, and governed by Eq. \eqref{eq:master}. For mutants with fitness greater than the population mean fitness, this process is known as ``establishment'', and occurs when $n_j$ is of order $1/s$, where $s$ is the mutant's fitness advantage relative to the mean \citep{desai_2007}. Here we do not consider the initial stochastic behavior of novel mutants, and have restricted our attention to the earliest deterministic behavior of rare genotypes. In particular, for beneficial mutations we have only considered the case where $s$ is large enough that deterministic behavior starts when $n_j \ll N$.

@@

We now discuss which changes in the traits $b, c$ and $d$ will be most favored under different environmental conditions. Of particular interest are Grime's ``disturbance'', ``stress'' and ``ideal'' environmental archetypes. To proceed, we need to map these verbal archetypes to quantitative parameter regimes in our model.

The ideal environmental archetype is characterized by the near-absence of stress and disturbance. Consequently, $d_i\ll 1$, whereas $b_i$ is potentially much larger than $1$. From Eq. \eqref{eq:deltN}, the equilibrium value of $L$ only depends on the ratio of birth and death rates. For one genotype, $L/(1-e^{-L})=b_i/d_i$, and so the propagule density is high $L\approx b_i/d_i\gg 1$. Moreover, since $L=b_i N/(N-T)$ by definition, population density is also high $N/T\approx 1$. Thus, territorial contests are decisively important.

The disturbance archetype is characterized by unavoidably high extrinsic mortality caused by physical destruction. Disturbances do not only affect adults as in Eq. \eqref{eq:delttot}, but also juveniles in the process of territorial contest. These juvenile deaths can be represented as a fractional reduction in the number of territories secured. To illustrate, we assume that the disturbance is equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by genotype $i$ each iteration. Then, the disturbance archetype is characterized by $d_i$ being close to $1$ for all genotypes (almost all adults and juveniles are killed each iteration). The single genotype equilibrium then gives $L\approx 2(1-d_i/[(1-d_i)b_i])$, where $b_i$ must be exceptionally large to ensure population persistence, and we have $L\ll 1$ and $N/T\ll 1$. The terms proportional to $c_i/\overline{c}$ in Eq. \eqref{eq:master} are then negligible, and $\Delta_+ n_i$ depends primarily on $b_i$.

@@

As noted in the previous section, if beneficial mutations establish (i.e survive the low-abundance stochastic regime), they will proceed to grow deterministically according to Eq. \eqref{eq:master}. The probability of establishment increases with the mutant fitness advantage, and is therefore typically on the order of one percent, whereas the fixation of neutral mutations is exceedingly unlikely (probability of order $1/N$). Consequently, the direction of evolutionary change is determined by which trait changes confer an appreciable benefit, subject to the constraints imposed by the environment.

@@

How does Fig.~4 compare to Grime's C/S/R strategies? In our comparison we will stick to fishes, corals and plants, for which three-way primary strategy schemes are well developed \citep{grime_1977,winemiller_1992,darling_2012}. The connection of our model to fish strategies is necessarily more tentative, given that fishes are motile and not all territorial and the starting assumption of our model is site-occupancy .

@@

Unlike Grime's classic ternary plot \citep{grime_1974}, which represents anti-correlations between traits relevant for success in different environmental archetypes, our realization of Grime's triangle (Fig.~4) refers instead to the direction of trait evolution under different regimes of stress and disturbance. As discussed in section ``Primary strategies and Grime's triangle'', over time our predicted trait evolution should lead to trait values consistent with Grime's scheme. In making these predictions, we have made no reference to any kind of trade-offs or pleitropy, even though trade-offs are often invoked to explain primary strategy schemes \citep{macarthur_1962,winemiller_1992,aerts_1999}. Thus, while trade-offs may amplify specialization, they are not necessary for it. For example, corals which rapidly out-shade neighbors have a tall, branched morphology which is vulnerable to disturbances, and so, all else being equal, ideal environment $c$-strategists will suffer higher mortality from disturbances. Fig.~\ref{fig:axes} gives the same conclusion without invoking trade-offs; mutations which reduce disturbance vulnerability are essentially neutral under ideal conditions, leading to no improvements in mortality from disturbances, whereas $c$ will tend to increase over time.

Our prediction of evolution for higher $b$ in ideal environments is counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1962} since $b$ is closely related to the maximal, low-density growth rate $b-d$, and ideal environments support high population densities which should be subject to ``$K$-selection''. However, in the Introduction, we noted that the $r$-$K$ dichotomy is not consistent with empirical studies showing that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}. From the perspective of our model, these correlations are not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$. Our higher-$b$ prediction simply reflects the fact that, all else being equal, producing more propagules is always advantageous, regardless of population density, a fact lost in the simple logistic interpretation of the $r$/$K$ scheme.

>>>>

Evolutionary models differ greatly in their treatment of fitness. In models of genetic evolution, genotypes are typically assigned constant (or frequency-dependent) selection coefficients describing the change in their relative frequencies over time due to differences in viability. This considerably simplifies the mathematics of selection, facilitating greater genetic realism, and can be justified over sufficiently short time intervals \citep[p. 276]{ewens_2012}.

However, a genotype's selective advantage

 can have very different effects when operating on different types of traits, and evolutionary changes in population can lead to complicated ecological responses.

@@

Far less work has been done to generalize beyond particular traits or ecological scenarios to models of fitness that still capture key distinctions between different forms of selection. Perhaps this is not surprising given that fitness is such a complex quantity, dependent on all of a phenotype's functional traits \citep{violle_2007} as well as its biotic and abiotic environment. In most cases, a detailed, trait-based, predictive model of fitness would be enormously complicated and have narrow applicability. It is therefore easy to doubt the feasibility of a simplified, general mathematical treatment of fitness \citep[p. 276]{ewens_2012}. For example, MacArthur's famous r/K selection scheme is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}. The r/K scheme's original purpose was as an extension of the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}, but few attempts have been made to develop it further as a mathematical analysis of the major different forms of selection.

@@

Trait classification schemes show empirically that, beneath the complicated details of trait variation, even among closely-related species, fitness is predominantly determined by a few key factors such as intrinsic reproducive rate or stress-tolerance. However, while trait classification schemes are firmly grounded in trait data, they are verbal and descriptive rather than mathematical, a recognized hinderance to their broader applicability (e.g. \citealt{tilman_2007}).

@@

We broadly follow the spirit of MacArthur's $r$/$K$ selection scheme in that our model is intended to account for fundamentally different forms of selection without getting  entangled in the intricacies of particular ecological scenarios. However, rather than building directly on MacArthur's formalism and its later extensions using Lotka-Volterra equations to incorporate competition (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001}, our model is devised primarily with Grime's C/S/R scheme in mind, and represents a quantitative formalization of how C/S/R manifests at the level of within-population genotype evolution (as opposed to phenotypic divergence between species). This choice is motivated in part by the substantial empirical support for the C/S/R scheme, and in part by the failings of the $r$/$K$ low/high density dichotomy --- many growth ability traits will confer advantages at both low and high densities (more details in the Discussion).

In the section ``Model'' we introduce the basic assumptions of our generalized lottery model. Analytical expressions for the change in genotype abundances over time are introduced in section ``Mean field approximation'', with mathematical details relegated to the Appendices. The following two sections discuss the behaviour of rare mutants and our treatment of Grime's triangle.

@@

Each iteration, adults produce $m_i$ new offspring (henceforth called ``propagules'') which disperse at random over the $U$ unnocupied territories (no dispersal limitation). We assume adults cannot be ousted from occupied territories, so only propagules landing on occupied territories are included in $m_i$. More generally, $m_i$ only includes propagules which actually end up contesting unnocupied territories. For simplicity, we assume $m_i=b_i n_i$, where $b_i$ is a constant, genotype-specific birth rate.

The number of individuals of the $i$'th genotype landing in any particular territory is denoted $x_i$. Random dispersal implies that in the limit $T\rightarrow \infty$, with $n_i/T$ held fixed, $x_i$ is Poisson distributed with mean territorial propagule density $l_i=m_i/U$. Although $T$ is finite in our model, we assume that $T$ and the $n_i$ are large enough that $x_i$ is Poisson-distributed to a good approximation (details in Appendix A). This dispersal Poisson distribution is denoted $p_i(x_i)$. Note that the large $n_i$, large $T$ approximation places no restrictions on our densities $n_i/T$, but it does preclude consideration of demographic stochasticity when $n_i$ itself is very small (this will be discussed further in Section ``Invasion of rare genotypes and coexistence'').

@@

The increase in $n_i$ over one iteration due to territorial acquisition, $\Delta_+ n_i$, is the sum of genotype $i$'s victories over all $U$ unoccupied territories. Since $p_1(x_1)\ldots p_G(x_G)$ is equal to the proportion of unoccupied territories with $x_1,\ldots,x_G$ of the respective propagules (where $G$ is the number of genotypes present; and again, we assume that $T$ is large enough that fluctuations in this proportion are negligible), this sum can be replaced by an expectation over the $p_i$. This gives
\begin{equation}
\Delta_+ n_i(t)=U(t)\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G). \label{eq:growthsumuncoupled}
\end{equation}

In addition to propagule birth and competition, occupied territories become unoccupied due to mortality. For the majority of this manuscript we assume that mortality only occurs in adults, and at a constant, genotype-specific per-capita rate $d_i$, so that the overall change in genotype abundances is
\begin{equation}
\Delta n_i(t)=\Delta_+ n_i(t)-d_i n_i(t). \label{eq:delttot}
\end{equation}
We will introduce a different mortality model when we consider the effects of disturbances (Section ``Primary strategies and Grime's triangle''), which will also affect competing juveniles.

Note that the competitive ability coefficients $c_i$ represent a strictly relative aspect of fitness in the sense that they only influence population size $N$ indirectly by changing genotype frequencies; that may in turn change the population mean birth and death rates. This can be seen by summing Eq. \eqref{eq:delttot} over genotypes to get the  change in population size $N$, 
\begin{equation}
\Delta N=U(1-e^{-L})-\sum_i d_i n_i,\label{eq:deltN}
\end{equation}
which is independent of $c_i$ (here $L=\sum_j l_j$ is the overalFaltinel propagule density).

@@

Our mean field approximation is similar to the high-$l_i$ approximation leading to Eq. \eqref{eq:lottery} in that we replace the $x_i$ with appropriate mean values. The key distinction is that territories with a single propagule from the focal genotype, whose behavior is critical at low densities, are handled separately. In place of the requirement of $l_i\gg 1$ for all $i$, our approximation only requires that there are no large discrepancies in competitive ability (specifically, that we do not have $c_i/c_j\gg 1$ for any two genotypes; further discussion in section ``Discussion''). We obtain (details in Appendix B)
\begin{equation}
\Delta_+ n_i(t)\approx b_i n_i\left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right], \label{eq:master}
\end{equation}
where
\begin{equation}
R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}\frac{\overline{c}L- c_il_i}{L-l_i}},\label{eq:Dr}
\end{equation}
and
\begin{equation}
A_i=\frac{\overline{c}(1-e^{-l_i})}{\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}c_il_i+\frac{1}{L-l_i}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right)\sum_{j\neq i}c_jl_j}.\label{eq:Da}
\end{equation}

Comparing Eq. \eqref{eq:master} to Eq. \eqref{eq:lottery}, the classic lottery per-propagule success rate $c_i/\overline{c}L$ has been replaced by three separate terms. The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these territories are won without contest. The second, $R_i c_i/\overline{c}$ represents competitive victories when the $i$ genotype is a rare invader in a high density population: from Eq. \eqref{eq:Dr}, $R_i\rightarrow 0$ when the $i$ genotype is abundant ($l_i\gg 1$), or other genotypes are collectively rare ($L-l_i\ll 1$). The third term, $A_ic_i/\overline{c}$, represents competitive victories when the $i$ genotype is abundant: $A_i\rightarrow 0$ if $l_i\ll 1$. The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $l_i/L$. If $l_i\gg 1$ for all genotypes, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$). Thus, Eq. \eqref{eq:master} generalizes the classic lottery model to account for arbitrary propagule densities for each genotype.

@@

\begin{figure}
\centering
\includegraphics[scale=0.7]{simulationcomparison.pdf}
\caption{\label{fig:simcomp} The change in genotype abundances in a density dependent lottery model is closely approximated by Eq. \eqref{eq:master}. $\Delta_+ n_i/m_i$ from Eq. \eqref{eq:master} (and its separate components) are shown, along with direct simulations of random dispersal and lottery competition over one iteration over a range of propagule densities (varied by changing $U$ with the $m_i$ fixed). Two genotypes are present. (a) and (b) show the low-frequency genotype with $c$-advantage ($m_1/M=0.1$, $c_1=1.5$), (c) and (d) show the high-frequency predominant genotype ($m_2/M=0.9$, $c_2=1$). Simulation points are almost invisible in (c) and (d) due to near exact agreement with Eq. \eqref{eq:master}.} 
\end{figure}

@@

In our model (section ``Model''), each genotype is defined by three traits: $b$, $c$ and $d$. To determine how these will evolve in a population where they are being modified by mutations, we need to know whether mutant lineages will grow (or decline) starting from low densities. In this section we discuss the behavior of rare genotypes predicted by Eq. \eqref{eq:master}.

Suppose that a population with a single genotype $i$ is in equilibrium. Then $R_i=0$, $\overline{c}=c_i$ and $\Delta n_i = 0$, and so Eq. \eqref{eq:master} gives
\begin{equation}
b_i\left(e^{-L}+A_i\right)-d_i=0,\label{eq:equil}
\end{equation}
where $A_i=(1-(1+L)e^{-L})/L$. Now suppose that a new genotype $j$, which is initially rare, appears in the population. Then $A_j\ll R_j$, $l_j\approx 0$ and $\overline{c}\approx c_i$, and so, from Eq. \eqref{eq:master}, $n_j$ will increase if 
\begin{equation}
b_j \left(e^{-L}+R_j\frac{c_j}{c_i}\right)-d_j>0,\label{eq:invad}
\end{equation}
where $R_j\approx (1-e^{-L})/\left(\frac{c_j}{c_i}+\frac{L-1-e^{-L}}{1-(1+L)e^{-L}}\right)$.

Combining Eqs. \eqref{eq:equil} and \eqref{eq:invad}, we see that $j$ will invade if it is superior in any one of the three traits, but is otherwise identical to $i$. If the new genotype has the same competitive ability $c_j=c_i$, then $R_j\approx A_i$ and Eqs. \eqref{eq:equil} and \eqref{eq:invad} imply that invasion occurs when $b_jd_i-b_id_j>0$, and in particular when $b_j>b_i$ with $d_j=d_i$, or when $d_j<d_i$ with $b_j=b_i$. In the case that the new genotype has a different competitive ability, Eqs. \eqref{eq:equil} and \eqref{eq:invad} imply that invasion occurs when $R_j c_j/c_i > A_i$; it is not hard to verify that this occurs if and only if $c_j>c_i$ using the simplified expressions for $A_i$ and $R_j$ given after Eqs. \eqref{eq:equil} and \eqref{eq:invad} respectively. Moreover, if $j$ invades in any of these cases, it will eventually exclude $i$, since it is strictly superior.

However, stable coexistence is possible between genotypes that are superior in different traits. To illustrate, suppose that $j$ is better at securing territories ($c_j>c_i$), that $i$ is better at producing propagules ($b_i>b_j$), and that $d_i=d_j$. Coexistence occurs if $j$ will invade an $i$-dominated population, but $i$ will also invade a $j$-dominated population (``mutual invasion''). It is not hard to show that this is possible, since if $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant, then, combining Eqs. \eqref{eq:equil} and \eqref{eq:invad}, we find that $i$ invades $j$ because $b_i>b_j$, while $j$ invades $i$ provided that
\begin{equation}
b_jc_jR_j-b_i c_i A_i>0. \label{eq:jinvadcoex}
\end{equation}
Thus, coexistence occurs if $c_j/c_i$ is large enough. Intuitively, the mechanism for coexistence is that territorial contests are important in an $i$-dominated population (high $L$), ensuring that the $c$-specialist $j$ is not excluded, yet territorial contests are irrelevant in a $j$-dominated population (low $L$), ensuring that the $b$-specialist $i$ is not excluded. Fig.~\ref{fig:coex} shows an example of this coexistence between $b$ and $c$ specialists.

@@

If the rare genotype $j$ arises due to mutation, then its initial low-density behavior is more complicated than the above invasion analysis suggests. The mutant lineage starts with one individual $n_j=1$, and remains at low abundance for many generations after its initial appearance. During this period, the mutant abundance $n_j$ will behave stochastically, and the deterministic equations \eqref{eq:growthsumuncoupled} and \eqref{eq:master} do not apply (section ``Model''). However, if $n_j$ becomes large enough, its behavior will become effectively deterministic, and governed by Eq. \eqref{eq:master}. For mutants with fitness greater than the population mean fitness, this process is known as ``establishment'', and occurs when $n_j$ is of order $1/s$ \citep{desai_2007}, where the selection coefficient $s$ is the mutant's fitness advantage (i.e. the mutant's fitness minus the population mean fitness). Here we do not consider the initial stochastic behavior of novel mutants, and have restricted our attention to the earliest deterministic behavior of rare genotypes. In particular, for beneficial mutations we have only considered the case where $s$ is large enough that deterministic behavior starts when $n_j \ll N$.

@@

We now discuss which changes in the traits $b, c$ and $d$ will be particularly favored under different environmental conditions. Of specific interest are Grime's ``disturbance'', ``stress'' and ``ideal'' environmental archetypes. To proceed, we need to map these verbal archetypes to quantitative parameter regimes in our model.

The ideal environmental archetype is characterized by the near-absence of stress and disturbance. Consequently, $d_i\ll 1$, whereas $b_i$ is potentially much larger than $1$. From Eq. \eqref{eq:deltN}, the equilibrium value of $L$ only depends on the ratio of birth and death rates. For one genotype, $L/(1-e^{-L})=b_i/d_i$, and so the propagule density is high $L\approx b_i/d_i\gg 1$. Moreover, since $L=b_i\frac{N}{N-T}=b_i\frac{1}{1-T/N}$ by definition, population density is also high $N/T\approx 1$. Thus, almost every unoccupied territory will be heavily contested.

The disturbance archetype is characterized by unavoidably high extrinsic mortality caused by physical destruction. Disturbances do not only affect adults as in Eq. \eqref{eq:delttot}, but also juveniles in the process of territorial contest. These juvenile deaths can be represented as a fractional reduction in the number of territories secured. To illustrate, we assume that the disturbance is equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by genotype $i$ each iteration. Then, the disturbance archetype is characterized by $d_i$ being close to $1$ for all genotypes (almost all adults and juveniles are killed each iteration). From Eq. \eqref{eq:deltN}, the single genotype equilibrium is given by $L/(1-e^{-L})=d_i/[(1-d_i)b_i]$, and since $L\ll 1$ and $N/T\ll 1$ due to high mortality, we have $L\approx 2(1-d_i/[(1-d_i)b_i])$. Clearly $b_i$ must be exceptionally large to ensure population persistence. The terms proportional to $c_i/\overline{c}$ in Eq. \eqref{eq:master} are then negligible, and $\Delta_+ n_i$ depends primarily on $b_i$.

@@

As noted in the previous section, if beneficial mutations establish (i.e survive the low-abundance stochastic regime), they will proceed to grow deterministically according to Eq. \eqref{eq:master}. The probability of establishment increases with the mutant fitness advantage, and is therefore typically on the order of one percent, whereas the fixation of neutral mutations is exceedingly unlikely (probability of order $1/N$). Consequently, the direction of evolutionary change is determined by which trait changes are both available, and confer an appreciable benefit, where availability is subject to constraints imposed by the environment.

@@

How does Fig.~4 compare to empirical analyses of Grime's C/S/R strategies? In our comparison we will stick to fishes, corals and plants, for which three-way primary strategy schemes are well developed \citep{grime_1977,winemiller_1992,darling_2012}. The connection of our model to fish strategies is necessarily more tentative, given that fishes are motile and not all territorial, and the starting assumption of our model is site-occupancy.

@@

Unlike Grime's classic ternary plot \citep{grime_1974}, which represents anti-correlations between traits relevant for success in different environmental archetypes, our realization of Grime's triangle (Fig.~4) refers instead to the direction of adaptive trait evolution under different regimes of stress and disturbance. As discussed in section ``Primary strategies and Grime's triangle'', over time our predicted trait evolution should lead to trait values consistent with Grime's scheme. In making these predictions, we have made no reference to any kind of trade-offs or pleitropy, even though trade-offs are often invoked to explain primary strategy schemes \citep{macarthur_1962,winemiller_1992,aerts_1999}. Thus, while trade-offs may amplify specialization, they are not necessary for it. As an example of a trade-off, corals which rapidly out-shade neighbors have a tall, branched morphology which is vulnerable to disturbances, and so, all else being equal, ideal environment $c$-strategists will suffer higher mortality from disturbances. Fig.~\ref{fig:axes} gives the same conclusion without invoking trade-offs; mutations which reduce disturbance vulnerability are essentially neutral under ideal conditions, leading to no improvements in mortality from disturbances, whereas $c$ will tend to increase over time.

Our prediction of evolution for higher $b$ in ideal environments is counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1962} since $b$ is closely related to the maximal, low-density growth rate $r=b-d$, and ideal environments support high population densities which should be subject to ``$K$-selection''. However, in the Introduction, we noted that the $r$-$K$ dichotomy is not consistent with empirical studies showing that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}. From the perspective of our model, these correlations are not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$. Our higher-$b$ prediction simply reflects the fact that, all else being equal, producing more propagules is always advantageous, regardless of population density, a fact lost in the simple logistic interpretation of the $r$/$K$ scheme.
